---
title: "R for Beginners: subetting data frames"
author: "C. Tobin Magle"
date: "Jan 11, 2017"
output: html_document
geometry: margin=0.5in
fontsize: 12pt
---
This material was adapted from the Software Carpentry [Inflammation](http://swcarpentry.github.io/r-novice-inflammation/01-starting-with-data.html) and [Gapminder]("https://raw.githubusercontent.com/swcarpentry/r-novice-gapminder/gh-pages/_episodes_rmd/data/gapminder-FiveYearData.csv") lessons. This document and associated script can be found at
[https://github.com/maglet/r-for-beginners](https://github.com/maglet/r-for-beginners)

```{r echo=FALSE}
gapminder <- read.csv("gapminder.csv") #reads data from the csv file
```

####Subsetting
Especially with large data sets, sometimes you need to assess a particular part 
of the data. Depending on what you want to do, you'll either want to work with 
a subset of rows or columns. We'll look at how to subset by both. 

#####Subsetting Columns and Rows
Let's say that you're only interested in the population data. Use the 
following code to extract only the population data and store it in a variable called
__pop__

```{r}
pop<-gapminder$pop
```

* You already know about the assignment operator (<-)
* The __$__ operator lets you subset the data fram based on column name
* __Syntax__: (variable name)__$__(column name)

Now you should see a new variable called __pop__  with 1704 
elements in your global environment. 

You can get the same subset using the "[row,column]" format:
```{r}
pop2<-gapminder[,3]
```

* __Syntax__: the row you want and the column you want, in brackets, separated 
by a comma
* Leaving either field blank, will select ALL elements, in this case all rows
* the population data are stored in column 32 of the data frame.

Prove to yourself that these data sets are equal using a __Boolean operator__ 
"==". Make sure to use the double equal sign when you're evaluating whether 
2 values are equivalent. A single "=" will try to perform an assignment operation.

Running

```{r eval=FALSE}
pop == pop2
```

returns a list of TRUE/FALSE values that indicate whether the values of each element in the vector match. Scanning through that list could get tedious,
so instead we can use the fact that __TRUE__ in R is equivalent to __1__, and 
__FALSE__ is the equivalent of __0__. Thus, if all elements are equal to each other,
the sum of that list of __TRUE__ s and __FALSE__ s would be equal to the number of 
rows.

```{r}
nrow(gapminder)==sum(pop == pop2)
```

It checks out! And you only have to look at one result.

Now let's calculate the mean of this column:

```{r}
mean(pop)
```

But what if you want to find the mean for all of the numeric columns? That's where the 
__apply__ function comes in. This function applies a function (__FUN__) to a 
chunk of data (__X__) on either the rows or columns (__MARGIN__).

Here's how you get the mean and standard deviation for all of the numeric columns:
```{r}
colmeans<-apply(X=gapminder[,c(3,5,6)],  #the data you want to use
                MARGIN=2,               #apply the function to the columns
                FUN=mean)               #take the mean of the values

colsds<-apply(X=gapminder[,c(3,5,6)],     #the data you want to use
             MARGIN=2,                  #apply the function to the colummns
             FUN=sd)                    #take the standard deviation of the columns
```

* __X=gapminder[,c(3,5,6)]__: grabs the 3rd to 32nd columns of the data frame
* __MARGIN = 2__: tells apply to run the function on each column (MARGIN=1 would
indicate rows)
* __FUN= mean__: tells apply to run the mean function 

You can get the mean for each patient as well by changing the __margin__.

```{r}
rowmeans<-apply(X=gapminder[,c(3,5,6)],    #the data you want to use
               MARGIN=1,                 #apply the function to rows
               FUN=mean)                 #take the mean of the rows
```

#####Subsetting by values

But what if you only want to see values for a certain country or continent? That's
where the __subset__ function comes in. Suppose you want to get population 
values only for African countries and store that subset in a variable called 
__AfricaPop__.

```{r}
AfricaPop <- subset(x= gapminder,            #the data frame to subset
                        subset = (continent == "Africa"))   #the factor to subset on
```

* __x = gapminder__: tells subset the data set we want to use
* __subset = (continent == "Africa")__: a boolean expression that tells subset to only 
select observations (rows) with the value "Africa" in the continent variable (column).

Let's do the same thing for Asian countries
```{r}
AsiaPop <- subset(x= gapminder,            #the data frame to subset
                        subset = (continent == "Asia"))   #the factor to subset on
```

Now we can compare the data for the two groups using a t-test:

```{r}
t.test(AfricaPop$pop, AsiaPop$pop)
```

That's a pretty low p value. 

Start accordion
<div id="accordion-start"></div>
Inside accordion
<div id="accordion-stop"></div>