---
title: "R-for-Beginners-Challenges"
output: html_document
---

```{r echo=FALSE}
gapminder <- read.csv("gapminder.csv") #reads data from the csv file
cat<-3
dog<-2
pony<-10000
```

## Setup

####1. What does the bottom right window (console) of the R studio interface do?
Runs code

####2. Why should you set up a .git repository in your project directory?
keep track of changes, backup online

####3. IF you don't specify otherwise, which directory is R studio workingin by default when you open a project

The project directory

## Operators

####1. How would you assign the value 467 to the variable kitten?

```{r}
kitten<-467
kitten=465
467->kitten
```
####2. How woud you determine if the variable cat is greater than the variable dog and that cat is less than pony??

```{r}
cat>dog & cat<pony
```

####3. Pretend you want to know whether the sum of red and blue is greater than the difference between green and purple. How would you write that out?

```{r eval = FALSE}
(red+blue)<(green-purple)
```
## Using functions
####1. How would you find the help file for par: a function that alters graphics parameters?
```{r eval = FALSE}
?par
help(par)
```
####2. What is the second argument of the t.test function?
y: an optional (non-empty) numeric vector of data values.

####3. What happens if you don't use the destfile argument in the download.file function?

Error message saying it's not optional

## Subsetting
### Challenge 1: Fix each of the following common data frame subsetting errors:

####1. Extract observations collected for the year 1957

```{r eval=FALSE}
# gapminder[gapminder$year = 1957,]  
gapminder[gapminder$year == 1957,]
```
####2. Extract all columns except 1 through to 4
```{r eval=FALSE}
# gapminder[,-1:4]  
gapminder[,-c(1:4)]
```
####3. Extract the rows where the life expectancy is longer the 80 years
```{r eval=FALSE}
# gapminder[gapminder$lifeExp > 80]  
gapminder[gapminder$lifeExp > 80,]
```
####4. Extract the first row, and the fourth and fifth columns (lifeExp and gdpPercap).
```{r eval=FALSE}
# gapminder[1, 4, 5]  
gapminder[1, c(4, 5)]
```
####5. Advanced: extract rows that contain information for the years 2002 and 2007
```{r eval=FALSE}
# gapminder[gapminder$year == 2002 | 2007,]   
gapminder[gapminder$year == 2002 | gapminder$year == 2007,]  
gapminder[gapminder$year %in% c(2002, 2007),]
```
##Plotting

Consider the scatter plot of GDP vs. life expectancy.
####1. How would you include only data from 2007? (remember subsetting)
```{r}
gapminder2007<-subset(gapminder,gapminder$year == 2007)
plot(gapminder2007$gdpPercap, gapminder2007$lifeExp,
     xlab="GDP per capita",        #X axis label
     ylab= "LifeExpectancy (years)",    #Y axis label (somehow get )
     las= 1,                  #Make numbers horizontal on the x axis, 
     cex.lab=1.2,             #Increase axis labels font size
     cex.axis=1.2,            #Increase axis label number size
     lwd=2,                   #Increase line/dot thickness
     bty = "n",               #removes black border around the plot
     pch=16,                  #Specifies symbols to draw
     ylim= c(0,100),            #set y axis limits
     col = gapminder$continent) #Set Color based on Continent

legend(x= 30000,                            #x position of legend
       y=50,                                #y position of legend
       legend=levels(gapminder$continent),  #Legend text 
       col = c(1:5),
       pch=16)                              #specifies how to draw symbols
```

####2. How would you change the data to log axis?
```{r}
gapminder2007<-subset(gapminder,gapminder$year == 2007)
plot(gapminder2007$gdpPercap, gapminder2007$lifeExp,
     log = "x",
     xlab="GDP per capita",        #X axis label
     ylab= "LifeExpectancy (years)",    #Y axis label (somehow get )
     las= 1,                  #Make numbers horizontal on the x axis, 
     cex.lab=1.2,             #Increase axis labels font size
     cex.axis=1.2,            #Increase axis label number size
     lwd=2,                   #Increase line/dot thickness
     bty = "n",               #removes black border around the plot
     pch=16,                  #Specifies symbols to draw
     ylim= c(0,100),            #set y axis limits
     col = gapminder$continent) #Set Color based on Continent

legend(x= 10000,                            #x position of legend
       y=40,                                #y position of legend
       legend=levels(gapminder$continent),  #Legend text 
       col = c(1:5),
       pch=16)                              #specifies how to draw symbols
```

####3. How would you make a similar plot from European countries, and colorize by year?

```{r}
gapminderEu<-subset(gapminder,gapminder$continent == "Europe")
plot(gapminderEu$gdpPercap, gapminderEu$lifeExp,
     log = "x",
     xlab="GDP per capita",        #X axis label
     ylab= "Life Expectancy (years)",    #Y axis label (somehow get )
     las= 1,                  #Make numbers horizontal on the x axis, 
     cex.lab=1.2,             #Increase axis labels font size
     cex.axis=1.2,            #Increase axis label number size
     lwd=2,                   #Increase line/dot thickness
     bty = "n",               #removes black border around the plot
     pch=16,                  #Specifies symbols to draw
     ylim= c(0,85),            #set y axis limits
     col = gapminder$year) #Set Color based on Continent

legend(x= 10000,                            #x position of legend
       y=62,                                #y position of legend
       legend=levels(as.factor(gapminder$year)),  #Legend text 
       col = c(1:5),
       pch=16)                              #specifies how to draw symbols
```

####4. Make a red histogram of population for all countries in 2007
```{r}
hist(x=gapminder$pop,    #plots life expectancy for all countries in all years
     main="Histogram of population in 2007", #main title
     xlab="Population",                        #x axis label
     ylab="Frequency",                                 #y axis label
     col="red")                                       #color
```

####5.